---
title: Include a title here
author: Include your name here
format: 
  beamer:
    theme: AnnArbor
    colortheme: dolphin
    fonttheme: structurebold
    link-citations: true
    linkcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
lang: en
bibliography: ../r_for_data_science_2_edition_umng.bib
csl: ../apa.csl
editor: visual
---

```{r r_libraries}
#| eval: true
#| echo: false
#| output: false
#| warning: false
#| error: false
# This is a r code chunk where you load libraries
# The name of the chunk is r_libraries
# The options specified as #| are execute options 
library(tidyverse)
library(kableExtra)
library(reticulate)
```

```{python py_modules}
#| eval: true
#| echo: false
#| output: false
#| warning: false
#| error: false
# This is a python code chunk where you load modules
# The name of the chunk is py_libraries
# The options specified as #| are execute options 
import numpy as np
```

# Contents

-   Heading 1
-   Heading 2
-   Heading 3
-   Heading 4
-   Heading 5
-   References

# Heading 1

-   According to bla, bla, bla [@wickham_r_2023]

    -   The citation use the file *r_for_data_science_2_edition_umng.bib*
    -   The citation applies the apa style with the file *apa.csl*
    
-   This is an **inline** r code

    -   My age is `r year(today(tzone = 'America/Bogota')) - year(ymd('1986-05-06',tz = 'America/Bogota'))` years

# Heading 2

- This is a table using `kableExtra`:

```{r table}
#| eval: true
#| echo: false
#| output: true
economics |>
  head(n=5) |>
  kbl(format = 'latex', booktabs = TRUE, 
      col.names = c('Date', 'Consumption', 'Population',
                    'Saving rate', 'Unemployment', 'Unemployed'),
      caption = 'US economic time series') |>
  kable_styling(latex_options = 'striped', font_size = 7) |>
  row_spec(row=0, bold=TRUE) |>
  column_spec(1, bold = TRUE, color = "red")
```

# Heading 3

-   This is a simple plot using `ggplot2`:

```{r plot}
#| out-width: '70%'
#| fig-align: 'center'
economics |>
  ggplot() + 
  geom_line(aes(x=date, y=pce)) + 
  labs(x='Year',
       y='Billions of dollars',
       title = 'Personal consumption expenditures')
```

# Heading 4

-   This is a R code chunk

```{r chunk}
#| echo: true
# Using R as a calculator
result <- 2 + 3
result
```

-   This is a Python code chunk

```{python}
#| echo: true
#  Creating a numpy array
np_array = np.array([*range(1,4)])
np_array
```

# Heading 5

-   From R to Python

```{python}
#| echo: true
np_array + r.result
```

-   From Python to R

```{r}
#| echo: true
py$np_array + result
```

# References {.allowframebreaks}
