---
title: Workflow basics
author: Luis Francisco Gomez Lopez
date: 2023-07-26
format: 
  beamer:
    theme: AnnArbor
    colortheme: dolphin
    fonttheme: structurebold
    link-citations: true
    linkcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
lang: en
bibliography: ../r_for_data_science_2_edition_umng.bib
csl: ../apa.csl
editor: visual
---

```{r r_libraries}
#| eval: true
#| echo: false
#| output: false
#| warning: false
#| error: false 
library(tidyverse)
library(mosaicCalc)
```

# Contents

-   Coding basics
-   Comments
-   What's in a name?
-   Calling functions
-   References

# Coding basics

-   R as a calculator

\tiny

```{r}
#| echo: true
#| eval: true
1 / 0
sqrt(2)^2
sin(pi)
```

# Coding basics

-   R as a calculator

    -   Calculus [@kaplan_mosaic_2022]

\small

$$\frac{dx^3}{dx} = 3x^2$$
\tiny

```{r}
#| echo: true
#| eval: true
mosaicCalc::D(x^3 ~ x)
```

\small

$$\int e^x dx = e^x + C$$
\tiny

```{r}
#| echo: true
#| eval: true
mosaicCalc::antiD(exp(x) ~ x)
```

# Coding basics

- Creating and **c**ombining objects

\tiny

```{r}
#| echo: true
#| eval: true
primes <- c(2, 3, 5, 7, 11, 13)
primes + 2
```

# Comments

-   Use comments to explain the **why** of your code, not the how or the what

    -   Don't describe every step in the comments

\tiny

```{r}
#| echo: true
#| eval: true
primes
# multiply primes by 3 to understand the concept of vectorization
primes * 3
```

# What's in a name?

-   Object names must start with a letter and can only contain letters, numbers, `_`, and `.`

\tiny

```{r}
#| echo: true
#| eval: true
#| error: true
1_primes <- c(2, 3, 5, 7, 11, 13)
```

\normalsize

-   Use **snake_case**

\tiny

```
i_use_snake_case
otherPeopleUseCamelCase
some.people.use.periods
And_aFew.People_RENOUNCEconvention
```

# What's in a name?

-   R is strict and case sensitive

\tiny

```{r}
#| echo: true
#| eval: true
#| error: true
my_variable <- 4
myvariable
My_variable
```

# Calling functions

- Function name, arguments and values

\tiny

```
function_name(argument1 = value1, argument2 = value2, ...)
```

\normalsize

-   An example with `seq`

\tiny

```{r}
#| echo: true
#| eval: true
seq(from = 2, to = 10, by = 2)
seq(by = 2, to = 10, from = 2)
# here we need to take into account the position of each argument
seq(2, 10, 2)
```

# References {.allowframebreaks}
