---
title: "Data transformation"
author: "Patricia Rodríguez-Sánchez"
date: 2023-08-07
institute: UMNG
lang: en
format: 
  beamer:
    number-sections: true 
editor: visual
---

```{r echo=TRUE}
library(tidyverse)

```

# Table of contents

4.1 Introduction

Functions that operate on rows and then columns of a data frame

The pipe, an important tool that you use to combine verbs.

The ability to work with groups.

Case study

# 4.1. Introduction

Theme: data transformation

Objective: learn key tools for transforming a data frame

Package of tidyverse: **dplyr**

Dataset: flights that departed from New York City in 2013

## 4.1.1 Package dplyr

```{r echo=TRUE}
library(nycflights13)
library(tidyverse)
```

```         
tidyverse_conflicts() ── 
dplyr::filter() masks stats::filter() 
dplyr::lag()    masks stats::lag() 

This means dplyr overwrites some functions in base R. 
We’ll use the same syntax as R: packagename::functionname().
```

## 4.1.2 **nycflights13**

To explore the basic dplyr verbs, we\'re going to use [`nycflights13::flights`](https://rdrr.io/pkg/nycflights13/man/flights.html)

```{r echo=TRUE}
nycflights13::flights

?flights
```

# References
